\scuba: iterative map-making with the Sub-Millimetre User Reduction Facility

### 2012-09-10 smurf_paper.tex
conclusions and acknowledgements

\begin{abstract}
  The Submillimetre Common User Bolometer Array 2 (\scuba) is an
  instrument operating on the 15-m James Clerk Maxwell Telescope,
  consisting of about 5000 bolometers in each of two simultaneous
  imaging bands centred over 450 and 850\,\micron. The camera is
  operated by scanning across the sky and records data at a rate of
  200\,Hz. This rate is much greater than that of previously existing
  submillimetre cameras, and represents a significant analysis
  challenge. We describe the production of \scuba\ maps using the
  Sub-Millimetre User Reduction Facility (SMURF) in which we have
  adopted a fast, iterative approach to map-making that enables data
  reduction on single, modern, high-end desktop computers, with
  execution times that are shorter than the observing times.  SMURF is
  used in an automated setting, both at the telescope for real-time
  feedback to observers, as well as for the production of generic
  science products at the Canadian Astronomy Data Centre. Three
  detailed case studies are used to: (i) explore convergence
  properties of the map-maker using simple prior constraints (Uranus
  -- a point source); (ii) achieve the white-noise limit for faint
  point-source studies (extragalactic blank-field survey of the
  Lockman Hole); and (iii) demonstrate that our strategy is capable of
  recovering angular scales comparable to the size of the array
  footprint (approximately 5\,arcmin) for bright extended sources
  (star-forming region M17).
\end{abstract}

@@

The third approach adopted here for \scuba\ is a compromise between
the previous two methods. Under the assumption that a significant
portion of the (predominantly low-frequency) non-white noise sources
can be modelled, an iterative solution is obtained for both the
astronomical image and the parameters of the noise model. Since the
remaining (non-modelled) noise sources are assumed to be white, a
single scalar \rms\ may be calculated for all of the data points from
a given bolometer to characterise its noise distribution (since the
noise at any instant in time is uncorrelated with others, and with the
data from other bolometers), greatly simplifying the measurement of
noise properties and the inversion step that is so complicated in the
maximum-likelihood methods. In principle, it should also be possible
to recover a wider dynamic range in angular scales than possible for
the simplest direct methods, with an increase in calculation time that
is merely linear in the number of iterations needed, while using a
comparable amount of memory. The iterative method of
\citet{wright1996} specific to chopped (CMB) data was applied to SCUBA
observations by \citet{johnstone2000}, although in this case the
chopping removes most of the low-frequency noise, and greatly
simplifies the problem (at the expensive of large-scale structure in
the data). More similar to the algorithm described here was the
iterative pipeline recipe for fitting and removing baseline drifts in
SCUBA scan-map data as the astronomical image estimate improved
\citep{1999ASPC..172..171J}. The two closest modern relatives of SMURF
are the Comprehensive Reduction Utility for SHARC-2
\citep[CRUSH,][]{kovacs2008}, and the pipeline developed for the
Bolocam Galactic Plane Survey \citep[][which uses iterative PCA
cleaning to avoid the suppression of large-scale
structure]{aguirre2010}.

A reasonable model for correlated noise in \scuba\ is a single
``common-mode'' signal seen by all of the bolometers. Iterative
estimation and removal of this signal significantly lowers the $1/f$
knee, without compromising structures on angular scales smaller than
the array footprint. Residual independent drifts at lower frequencies
are removed with an iterative high-pass filter. This strategy enables
SMURF to reduce data faster than they are taken, on single high-end
desktop computers, and it has successfully been used as part of a
real-time pipeline offering feedback to observers at the
telescope. The pipeline is also used to generate near-science grade
products using the JCMT Science Archive \citep{2011ASPC..442..203E}
hosted by the Canadian Astronomy Data Centre (CADC). Details of the
pipeline design are given in \citet{gibb2005} and
\citet{2008AN....329..295C}.

@@

Finally, we experimented with an alternative to \model{COM} for
removing correlated atmospheric noise. Rather than subtracting the
average signal at each time slice, \model{PLN} fits a plane to the
observed signal at each instant. We found that there was no obvious
improvement (either in terms of reducing the $1/f$ knee or reducing
the noise in final maps). This result is basically consistent with
those from earlier instruments, and suggests that the angular scale of
the emitting regions in the atmosphere are un-resolved at the \scuba\
focal plane. \citet{sayers2010}, in particular, examined several
different ways of modelling and removing the correlated atmospheric
noise from Bolocam data at 143 and 268\,GHz (also atop Mauna Kea),
finding only marginal improvements by fitting a plane, or even using
higher-order polynomials (see their Fig.~10, as well as the discussion
and references to earlier work in their Section~4.3). In
\citet{aguirre2010} iterative PCA cleaning was used to remove the
atmosphere since the simpler methods mentioned left substantial
residuals in their data (similar to the \scuba\ data described here,
although PCA cleaning is prohibitively slow in this case).

@@

Note that alternatives to the zero-masking approach do exist for other
iterative map-makers. For example, \citet{kovacs2008} typically
restricts the solution to a small fixed number of iterations
($\lsim$10), so that there is probably little chance that such
structures have the chance to grow. Experimentation with the model
order is also advocated to gain an impression of the convergence
properties. This approach is perhaps more relevant to their SHARC-2
data for which a more complicated model is developed; the degeneracies
we have observed for \scuba\ maps using our more brute-force approach
with a high-pass filter tend to look similar regardless of the order
(only the convergence time is affected).  For Bolocam data of the
Galactic Plane, \citet{aguirre2010} used a maximum-entropy filtering
step to suppress large-scales. Regardless of the method used,
simulations are always required to establish the transfer function and
noise properties as a function of angular scale.

@@

This paper has described the Submillimetre User Reduction Facility
(SMURF), which was designed to produce maps from the rapidly sampled
$\sim10^4$ bolometers of the \scuba\ instrument. We have presented
sufficient detail in order to be useful not only to users of \scuba,
but for similar detector arrays in the future. The basic elements are:

\item A simple model which removes most noise in a common-mode
rejection step, along with high-pass filtering, results in maps that
approach the white-noise performance of the detectors.

\item The main problem to overcome with our strategy is the divergence
of the map on large angular scales, due to the degeneracy between the
map, common-mode, and other residual low-frequency noise
contributions. A simple strategy of constraining empty regions of the
map to zero (either a user-supplied mask for known sources, or an
iterative determination of signal above some \snr\ threshold) provides
good constraints for compact objects, and bright/extended structures.

\item For maps of faint point sources, a single (non-iterative)
high-pass filter at the start of the reduction produces maps that are
nearly white-noise limited. We illustatre how to completely whiten the
maps using jackknife measurements of the angular noise power spectrum
in order to reliably extract sources down to the limits of the map.

\item The map-maker has an automatic stopping criterion based on the
change in the map and $\chi^2$ between subsequent iterations, allowing
it to adapt to a wide range of data in a pipeline setting.

\item Principal Component Analysis (PCA) has been used to identify
correlated noise contributions to \scuba\ data, and it is available as
a generic data cleaning option.

\item The memory requirements and execution times are reasonable
(possible to reduce data in less time than it takes to collect) on
modern high-end desktop computers due to the simplicity of the SMURF
algorithm. It is used in real-time at the telescope for feedback
during observations, and it is also regularly produces nearly
science-grade products at the Canadian Astronomy Data Centre.

\item SMURF is highly configurable, making it easy to vary the order
and types of models that are fit to the data. It is also easy to add
new models to the existing infrastructure.

\item It is believed that a maximum-likelihood map-maker, such as
SANEPIC, will provide improved maps over the iterative algorithm
described here. However, SMURF will most likely be used as an initial
step in such processing since it can quickly clean the bolometer
time-series, as well as perform map-based despiking (a necessarily
iterative step).

@@

The James Clerk Maxwell Telescope is operated by the Joint Astronomy
Centre on behalf of the Science and Technology Facilities Council of
the United Kingdom, the Netherlands Organisation for Scientific
Research, and the National Research Council of Canada. Additional
funds for the contruction of \scuba\ were provided by the Canada
Foundation for Innovation. EC thanks CANARIE/CANFAR for additional
funding. This research was supported in part by the Natural Sciences
and Engineering Research Council of Canada. The authors thank the
members of the SCUBA-2 commissioning team for testing the
map-maker and reporting anomalies; in particular Antonio Chrysostomou
and Jessica Dempsey. We also thank Dennis Kelly, Alex van Engelen and
Jennifer Balfour for work on the SCUBA-2 observation simulator;
and Mark Thompson, Craig Walther and S\'{e}verin Gaudet for being
on the Critical Design Review panel.

>>>>

\begin{abstract}
  The Submillimetre Common User Bolometer Array 2 (\scuba) is an
  instrument operating on the 15-m James Clerk Maxwell Telescope,
  consisting of about 5000 bolometers in each of two simultaneous
  imaging bands centred over 450 and 850\,\micron. The camera is
  operated by scanning across the sky and records data at a rate of
  200\,Hz. This rate is much greater than that of previously existing
  submillimetre cameras, and represents a significant analysis
  challenge. We describe the production of \scuba\ maps using the
  Sub-Millimetre User Reduction Facility (SMURF) in which we have
  adopted a fast, iterative approach to map-making that enables data
  reduction on single, modern, high-end desktop computers, with
  execution times that are shorter than the observing times.  SMURF is
  used in an automated setting, both at the telescope for real-time
  feedback to observers, as well as for the production of generic
  science products for the JCMT Science Archive at the Canadian
  Astronomy Data Centre. Three detailed case studies are used to: (i)
  explore convergence properties of the map-maker using simple prior
  constraints (Uranus -- a point source); (ii) achieve the white-noise
  limit for faint point-source studies (extragalactic blank-field
  survey of the Lockman Hole); and (iii) demonstrate that our strategy
  is capable of recovering angular scales comparable to the size of
  the array footprint (approximately 5\,arcmin) for bright extended
  sources (star-forming region M17).
\end{abstract}

@@

The third approach adopted here for \scuba\ is a compromise between
the previous two methods. Under the assumption that a significant
portion of the (predominantly low-frequency) non-white noise sources
can be modelled, an iterative solution is obtained for both the
astronomical image and the parameters of the noise model. Since the
remaining (non-modelled) noise sources are assumed to be white, a
single scalar \rms\ may be calculated for all of the data points from
a given bolometer to characterise its noise distribution (since the
noise at any instant in time is uncorrelated with others, and with the
data from other bolometers), greatly simplifying the measurement of
noise properties and the inversion step that is so complicated in the
maximum-likelihood methods. In principle, it should also be possible
to recover a wider dynamic range in angular scales than possible for
the simplest direct methods, with an increase in calculation time that
is merely linear in the number of iterations needed, while using a
comparable amount of memory. The iterative method of
\citet{wright1996} specific to chopped (CMB) data was applied to SCUBA
observations by \citet{johnstone2000}, although in this case the
chopping removes most of the low-frequency noise, and greatly
simplifies the problem (at the expensive of large-scale structure in
the data). More similar to the algorithm described here was the
iterative pipeline recipe for fitting and removing baseline drifts in
SCUBA scan-map data as the astronomical image estimate improved
\citep{1999ASPC..172..171J}. The two closest modern relatives of SMURF
are the Comprehensive Reduction Utility for SHARC-2
\citep[CRUSH,][]{kovacs2008}, and the pipeline developed for the
Bolocam Galactic Plane Survey \citep[][which uses iterative PCA
cleaning to avoid the suppression of large-scale
structure]{aguirre2011}.

A reasonable model for correlated noise in \scuba\ is a single
``common-mode'' signal seen by all of the bolometers. Iterative
estimation and removal of this signal significantly lowers the $1/f$
knee, without compromising structures on angular scales smaller than
the array footprint. Residual independent drifts at lower frequencies
are removed with an iterative high-pass filter. This strategy enables
SMURF to reduce data faster than they are taken, on single high-end
desktop computers, and it has successfully been used as part of a
real-time pipeline offering feedback to observers at the
telescope. The pipeline is also used to generate near-science grade
products for the JCMT Science Archive \citep{2011ASPC..442..203E}
hosted by the Canadian Astronomy Data Centre (CADC). Details of the
pipeline design are given in \citet{gibb2005} and
\citet{2008AN....329..295C}.

@@

Finally, we experimented with an alternative to \model{COM} for
removing correlated atmospheric noise. Rather than subtracting the
average signal at each time slice, \model{PLN} fits a plane to the
observed signal at each instant. We found that there was no obvious
improvement (either in terms of reducing the $1/f$ knee or reducing
the noise in final maps). This result is basically consistent with
those from earlier instruments, and suggests that the angular scale of
the emitting regions in the atmosphere are un-resolved at the \scuba\
focal plane. \citet{sayers2010}, in particular, examined several
different ways of modelling and removing the correlated atmospheric
noise from Bolocam data at 143 and 268\,GHz (also atop Mauna Kea),
finding only marginal improvements by fitting a plane, or even using
higher-order polynomials (see their Fig.~10, as well as the discussion
and references to earlier work in their Section~4.3). In
\citet{aguirre2011} iterative PCA cleaning was used to remove the
atmosphere since the simpler methods mentioned left substantial
residuals in their data (similar to the \scuba\ data described here,
although PCA cleaning is prohibitively slow in this case).

@@

Note that alternatives to the zero-masking approach do exist for other
iterative map-makers. For example, \citet{kovacs2008} typically
restricts the solution to a small fixed number of iterations
($\lsim$10), so that there is probably little chance that such
structures have the chance to grow. Experimentation with the model
order is also advocated to gain an impression of the convergence
properties. This approach is perhaps more relevant to their SHARC-2
data for which a more complicated model is developed; the degeneracies
we have observed for \scuba\ maps using our more brute-force approach
with a high-pass filter tend to look similar regardless of the order
(only the convergence time is affected).  For Bolocam data of the
Galactic Plane, \citet{aguirre2011} used a maximum-entropy filtering
step to suppress large-scales. Regardless of the method used,
simulations are always required to establish the transfer function and
noise properties as a function of angular scale.

@@

This paper has described the Submillimetre User Reduction Facility
(SMURF), which was designed to produce maps from the rapidly sampled
$\sim10^4$ bolometers of the \scuba\ instrument. While similar to
other algorithms in the literature that iteratively model and remove
correlated noise components from the data, successively improving the
map \citep[e.g.,][]{kovacs2008,aguirre2011}, we have shown that a
relatively simpler approach provides good results for \scuba\ data,
with reasonable computational requirements. This conclusion is
encouraging for the development of future large-format bolometer
arrays.

%We have presented
%sufficient detail in order to be useful not only to users of \scuba,
%but for similar detector arrays in the future. The basic elements are:

The most significant correlated noise sources in \scuba\ data occur at
low-frequencies ($\lsim$2\,Hz), and are probably a mixture of
atmospheric signals and magnetic field pickup. Most of this signal can
be removed using common-mode rejection, although a principal component
analysis (PCA) identifies a number of other less-significant
correlated noise patterns that also dominate at low
frequencies. Unfortunately these patterns are difficult to model, and
their character varies from observation to observation. One could use
the PCA directly to remove them, but this technique is prohibitively
slow for existing high-end desktop computers given the volume of
\scuba\ data. However, we have demonstrated that high-pass filtering
can remove the bulk of these residual correlated noise components
using significantly more efficient Fast Fourier Transforms (FFTs). The
basic algorithm therefore consists of iterative common-mode rejection
and high-pass filtering along with estimates of the map, which allows
us to approach the white-noise limit of the instrument.

The main problem to overcome is divergence on large angular scales,
due to the degeneracy between the map, and the low-frequency signal
components that are removed (model degeneracies are a generic problem
of iterative map-makers). A simple strategy of constraining empty
regions of the map to zero (using either a user-supplied mask for
known sources, or an iterative determination of signal above some
\snr\ threshold) provides good constraints for compact objects, and
bright/extended structures, allowing us to effectively recover angular
scales up to the order of the array footprint (approximately
5\,arcmin).

For maps of faint point sources, a single (non-iterative) high-pass
filter at the start of the reduction produces maps that are nearly
white-noise limited. Residual large-scale noise can be removed with a
whitening filter, and sources detected using a matched-filter
(smoothing with the effective filtered point spread function).

The iterative solution is stopped once convergence in the $\chi^2$
statistic, and/or variations in the map itself is achieved. This
enables SMURF to run in a pipeline setting without user interaction
for a wide variety of observations. Furthermore, the execution times
are shorter than the observation lengths, and memory requirements for
even the longest \scuba\ observations are within the capabilities of
single, high-end desktop computers. SMURF can therefore provide real-time
feedback at the telescope to observers, and it also produces near
science-grade data products for the JCMT Science Archive hosted by the
Canadian Astronomy Data Centre.

%SMURF is highly configurable, making it easy to vary the order
%and types of models that are fit to the data. It is also easy to add
%new models to the existing infrastructure.

One regime in which SMURF does not presently perform well is in maps
of faint extended structures, since the zero-masking technique we have
adopted cannot be used. Since SMURF is both highly configurable and
extensible, it may be possible to develop an improved data model
and/or map constraint to assist in these situations, as more
experience with the instrument is gained. However, the best long-term
prospects are probably to turn to a maximum-likelihood algorithm, such
as SANEPIC \citep{patanchon2008}.

@@

The James Clerk Maxwell Telescope is operated by the Joint Astronomy
Centre on behalf of the Science and Technology Facilities Council of
the United Kingdom, the Netherlands Organisation for Scientific
Research, and the National Research Council of Canada. Additional
funds for the contruction of \scuba\ were provided by the Canada
Foundation for Innovation. EC thanks CANARIE/CANFAR for additional
funding. This research was supported in part by the Natural Sciences
and Engineering Research Council of Canada. EC also thanks Matthew
Hasselfield and Mark Halpern for helpful discussions about multiplexed
TES arrays and their noise characteristics.  The authors thank the
members of the SCUBA-2 commissioning team for testing the map-maker
and reporting anomalies; in particular Antonio Chrysostomou and
Jessica Dempsey. We also thank Dennis Kelly, Alex van Engelen and
Jennifer Balfour for work on the SCUBA-2 observation simulator; and
Mark Thompson, Craig Walther and S\'{e}verin Gaudet for being on the
Critical Design Review panel. Finally, we thank Per Friberg and Gary
Davis for their helpful comments on the manuscript.
